A Coq Formalization of Stable Sorts, Radix Sort, and the Burrows-Wheeler Transform

### 2019-10-19 draft_paper/draft_paper.tex
Fix preimage/image issue

A permutation function on $n$ elements maps indices to indices, and so
has type $\nat_{<n} \to \nat_{<n}$. In Coq, we can represent a finite
function as a list of its outputs: $f : \nat_{<n} \to A$ is represented
by the list $[f(0),~f(1),\allowbreak\ldots,~f(n-1)]$. Applied to
permutations, this means that the permutation~$p$ mentioned above
would be represented by $[2,~1,~0]$. This is consistent with
``one-line'' notation for permutations, in which a permutation on $n$
elements is represented by its result when applied to the
list~$[0,~2,\ldots,n-1]$. Now we can define the action of a permutation on
a list.
\begin{equation*}
  p(l) \coloneq \var{map}~(\lambda i. l_i)~p
\end{equation*}
We implement this in Coq using an \var{apply} function. The definition
is slightly complicated by the fact that \var{nth} takes an extra
parameter to return in case $i$ is out of bounds.
\begin{lstlisting}
Definition apply p l : list A :=
  match l with
  | [] => []
  | d :: _ => map (fun i => nth i l d) p
  end.
\end{lstlisting}

@@

\begin{definition}[StablePermFun]
  A permutation function $p$ is a \textit{stable} permutation function
  on some list $l$ when for all indices $i < j$ where $l_i \equiv l_j$, we
  also have $p(i) < p(j)$.
\end{definition}

@@

Now, instead of indexing into each column at $i$, where the columns
are all the result of successively applying $p$ to $l$, we will
successively take the image of $i$ under $p$ to generate the indices
into $l$ where those characters appear.
\begin{lstlisting}
Definition image (p : list nat) i := nth i p 0.
\end{lstlisting}
This is justified via the following theorem.
\begin{theorem}
  For all $n$,
  \begin{equation*}
    \var{map}~(\lambda r.~r_i)~(\var{iter}~(\var{apply}~p)~n~l)
    =
    \var{map}~(\lambda j.~l_j)~(\var{iter}~(\var{image}~p)~n~i)
  \end{equation*}
\end{theorem}
\begin{proof}
  \begin{align*}
    & \var{map}~(\lambda r.~r_i)~(\var{iter}~(\var{apply}~p)~n~l) \\
    ={}& \var{map}~(\lambda r.~r_i)~[l,~(\var{apply}~p)~l,~(\var{apply}~p)^2~l,~\ldots,~(\var{apply}~p)^{n-1}~l] \\
    ={}& [l_i,~((\var{apply}~p)~l)_i,~((\var{apply}~p)^2~l)_i,~\ldots,~((\var{apply}~p)^{n-1}~l)_i] \\
    ={}& [l_i,~l_{p(i)},~l_{p^2(i)},~\ldots,~l_{p^{n-1}~i}] \\
    ={}& \var{map}~(\lambda j.~l_j)~[i,~p(i),~p^2(i),~\ldots,~p^{n-1}~i] \\
    ={}& \var{map}~(\lambda j.~l_j)~(\var{iter}~(\var{image}~p)~n~i) \qedhere
  \end{align*}
\end{proof}

@@

We can define the iamge and preimage of an index under a permutation
function.
\begin{lstlisting}
Definition image (p : list nat) i := nth i p 0.
Definition preimage (p : list nat) i := findIndex i p.
\end{lstlisting}
We will use the notation $p(i)$ for $\var{image}~i$ and $p^{-1}(i)$
for $\var{preimage}~i$. It should be clear from context whether $p(x)$
denotes $\var{image}~x$ or $\var{apply}~x$, and when it is not we will
use \var{image} and \var{apply} explicitlity.

We will not prove it here, but $\var{preimage}~p$ and $\var{image}~p$
are both of bijections of type $\nat_{<n} \to \nat_{<n}$ and are
inverses of each other. We also have the following two theorems:
\begin{theorem}[nth\_preimage\_apply, nth\_image\_apply]
  For all lists $l$, permutations $p$, and indices $i$,
  \begin{align*}
    (\var{apply}~p~l)_{p^{-1}(i)} = l_i && l_{p(i)} = (\var{apply}~p~l)_i
  \end{align*}
\end{theorem}
\begin{proof}
  The first equation (\var{nth\_preimage\_apply}) holds because
  \begin{align*}
       (\var{apply}~p~l)_{p^{-1}(i)}
    &= (\var{map}~(\lambda j.~l_j)~p)_{\var{findIndex}~i~p} \\
    &= l_{p_{(\var{findIndex}~i~p)}} \\
    &= l_i
  \end{align*}
  and the second follows from the first.
  \begin{equation*}
      (\var{apply}~p~l)_i
    = (\var{apply}~p~l)_{p^{-1}(p(i))}
    = l_{p(i)}
    \qedhere
  \end{equation*}
\end{proof}

@@

>>>>

One natural representation for a permutation function~$p$ is a list of
natural numbers, where each element~$i$ of $p$ stands in for the $i$th
element of the original list. For example, the representation of the
permutation mapping $l =
[\var{zebra},\ \allowbreak\var{appliance},\ \allowbreak\var{kite}]$ to
$l' =
[\var{appliance},\ \allowbreak\var{kite},\ \allowbreak\var{zebra}]$
would be $[1, 2, 0]$, because $\var{appliance} = l_1$, $\var{kite} =
l_2$ and $\var{zebra} = l_0$. So, the mathematical permutation
function~$p$ on $n$ elements is represented by the list
$[p^{-1}(0),\ \allowbreak p^{-1}(1),\ \allowbreak,\ldots,\ \allowbreak
  p^{-1}(n-1)]$. This gives us a very elegant definition for the
action of a permutation on lists.
\begin{equation*}
  p(l) \coloneq \var{map}~(\lambda i. l_i)~p
\end{equation*}
We implement this in Coq with \var{apply}. The definition is slightly
complicated by the fact that \var{nth} takes an extra parameter to
return in case $i$ is out of bounds.
\begin{lstlisting}
Definition apply p l : list A :=
  match l with
  | [] => []
  | d :: _ => map (fun i => nth i l d) p
  end.
\end{lstlisting}

@@

\begin{definition}[StablePermFun]
  \label{def:stablepermfun}
  A permutation function $p$ is a \textit{stable} permutation function
  on some list $l$ when for all indices $i < j$ where $l_i \equiv l_j$, we
  also have $p(i) < p(j)$.
\end{definition}

@@

Now, instead of indexing into each column at $i$, where the columns
are all the result of successively applying $p$ to $l$, we will
successively take the image of $i$ under $p$ to generate the indices
into $l$ where those characters appear.
\begin{lstlisting}
Definition image (p : list nat) i := findIndex i p.
\end{lstlisting}
This is justified via the following theorem.
\begin{theorem}
  For all $n$,
  \begin{equation*}
    \var{map}~(\lambda r.~r_i)~(\var{iter}~(\var{apply}~p)~n~l)
    =
    \var{map}~(\lambda j.~l_j)~(\var{iter}~(\var{image}~p)~n~i)
  \end{equation*}
\end{theorem}
\begin{proof}
  \begin{align*}
    & \var{map}~(\lambda r.~r_i)~(\var{iter}~(\var{apply}~p)~n~l) \\
    ={}& \var{map}~(\lambda r.~r_i)~[l,~(\var{apply}~p)~l,~(\var{apply}~p)^2~l,~\ldots,~(\var{apply}~p)^{n-1}~l] \\
    ={}& [l_i,~((\var{apply}~p)~l)_i,~((\var{apply}~p)^2~l)_i,~\ldots,~((\var{apply}~p)^{n-1}~l)_i] \\
    ={}& [l_i,~l_{p(i)},~l_{p^2(i)},~\ldots,~l_{p^{n-1}~i}] \\
    ={}& \var{map}~(\lambda j.~l_j)~[i,~p(i),~p^2(i),~\ldots,~p^{n-1}~i] \\
    ={}& \var{map}~(\lambda j.~l_j)~(\var{iter}~(\var{image}~p)~n~i) \qedhere
  \end{align*}
\end{proof}

@@

We can define the image and preimage of an index under a permutation
function.
\begin{lstlisting}
Definition image (p : list nat) i := findIndex i p.
Definition preimage (p : list nat) i := nth i p 0.
\end{lstlisting}
We will use the notation $p(i)$ for $\var{image}~i$ and $p^{-1}(i)$
for $\var{preimage}~i$. It should be clear from context whether $p(x)$
denotes $\var{image}~x$ or $\var{apply}~x$, and when it is not we will
use \var{image} and \var{apply} explicitlity.

We will not prove it here, but $\var{preimage}~p$ and $\var{image}~p$
are both of bijections of type $\nat_{<n} \to \nat_{<n}$ and are
inverses of each other. We also have the following two theorems:
\begin{theorem}[nth\_preimage\_apply, nth\_image\_apply]
  For all lists $l$, permutations $p$, and indices $i$,
  \begin{align*}
    (\var{apply}~p~l)_{p(i)} = l_i && l_{p^{-1}(i)} = (\var{apply}~p~l)_i
  \end{align*}
\end{theorem}
\begin{proof}
  The first equation (\var{nth\_preimage\_apply}) holds because
  \begin{align*}
       (\var{apply}~p~l)_{p(i)}
    &= (\var{map}~(\lambda j.~l_j)~p)_{\var{findIndex}~i~p} \\
    &= l_{p_{(\var{findIndex}~i~p)}} \\
    &= l_i
  \end{align*}
  and the second follows from the first.
  \begin{equation*}
      (\var{apply}~p~l)_i
    = (\var{apply}~p~l)_{p(p^{-1}(i))}
    = l_{p^{-1}(i)}
    \qedhere
  \end{equation*}
\end{proof}

@@

We defined \var{StablePermFun} in terms of the indices of the original
list~$l$ (\cref{def:stablepermfun}), but we just as easily could have
defined it in terms of the indices of the permuted list~$l'$.

\begin{definition}[StablePermFun\_preimage]
  A permutation function $p$ is a \textit{stable} permutation function
  on some list $l$ when for all indices $i < j$ where $p(l)_i \equiv p(l)_j$, we
  also have $p^{-1}(i) < p^{-1}(j)$.
\end{definition}

\begin{theorem}[StablePermFun\_iff]
  For all lists~$l$ and $p : \var{list}~\nat$,
  \begin{equation*}
    \var{StablePermFun}~l~p \iff \var{StablePermFun\_preimage}~l~p
  \end{equation*}
\end{theorem}
\begin{proof}
  Both directions are very similar, we will prove just the $\implies$
  direction here.

  We have that $\var{StablePermFun}~l~p$, and we have an $i$ and $j$
  such that $i < j$ and $p(l)_i \equiv p(l)_j$. We must show that
  $p^{-1}(i) < p^{-1}(j)$. For the sake of contradiction, suppose
  $p^{-1}(i) \ge p^{-1}(j)$. If $p^{-1}(i) = p^{-1}(j)$, then $i = j$,
  contradicting $i < j$, so suppose $p^{-1}(i) > p^{-1}(j)$. Then,
  since $p(l)_i \equiv p(l)_j$ implies $l_{p^{-1}(i)} \equiv l_{p^{-1}(j)}$,
  from $\var{StablePermFun}~l~p$ we know that $p(p^{-1}(i)) >
  p(p^{-1}(i))$, and therefore that $i > j$. This is also a
  contradiction, so it must be that $p^{-1}(i) < p^{-1}(j)$.
\end{proof}
